---
title: "DESeq2 - Differential Expression Analysis of N.truncata and N. excelsior roots under drought stress"
output: html_notebook
---

# Install & Load libraries
```{r}
#Load libraries
pacman::p_load("edgeR", "ggplot2", "ggpubr", "tidyverse", "gridExtra", "ggfortify", "pheatmap", "VennDiagram", "plotly", "ggrepel", "gprofiler2", "topGO", "dplyr", "xlsx", "tibble", "readxl", "statmod")
```

```{r}
library(DESeq2)
library(plotly)
library(apeglm)
```

# Data Preprocessing:
```{r}
#Load Data of STAR alignment and featureCount
data <- read.table("/Users/daviddornig/Documents/botany_project/data/TRU_EXC_Roots/TRU_EXC_Roots_fragmentcount.txt", header = TRUE)
colnames(data) <- gsub("Aligned.sortedByCoord.out.bam", "", colnames(data))
col_names <- colnames(data)
colnames(data) <- gsub("^X", "", col_names)
samplelist <- colnames(data)[-(1:6)]
print(samplelist)
```
## dataframe of STAR aligned and featureCount data:
```{r}
data
```

## Prepare metadata from experimental design excel file as meta_df
```{r}
metadata <- c(
  "61;excelsior;drought;roots",
  "62;excelsior;drought;roots",
  "63;excelsior;control;roots",
  "64;excelsior;drought;roots",
  "65;excelsior;control;roots",
  "66;excelsior;control;roots",
  "67;excelsior;control;roots",
  "68;excelsior;drought;roots",
  "69;excelsior;drought;roots",
  "70;excelsior;drought;roots",
  "71;excelsior;control;roots",
  "72;excelsior;control;roots",
  "85;truncata;control;roots",
  "86;truncata;drought;roots",
  "87;truncata;drought;roots",
  "88;truncata;control;roots",
  "89;truncata;drought;roots",
  "90;truncata;control;roots",
  "91;truncata;drought;roots",
  "92;truncata;control;roots",
  "93;truncata;control;roots",
  "94;truncata;drought;roots",
  "95;truncata;drought;roots",
  "96;truncata;control;roots"
)
```
# Create sample dataframe with metadata:
```{r}

# Define a function to generate abbreviations like ECR85 from excelsior, control, roots, 85
generate_abbreviation <- function(line) {
  # Split the line into components
  components <- strsplit(line, ";")[[1]]
  
  # Extract metadata components
  species <- components[2]
  condition <- components[3]
  location <- components[4]
  number <- components[1]
  
  # Combine components to generate abbreviation
  abbreviation <- paste0(toupper(substr(species, 1, 1)), toupper(substr(condition, 1, 1)), toupper(substr(location, 1, 1)), number)
  
  return(abbreviation)
}

abbreviations <- sapply(metadata, generate_abbreviation)


# Split the metadata into ID, Species, and Condition for df conversion
metadata_split <- strsplit(metadata, ";")
split_abbreviations <- strsplit(abbreviations, ";")


ID = sapply(metadata_split, `[`, 1)
Species = sapply(metadata_split, `[`, 2)
Condition = sapply(metadata_split, `[`, 3)
Location = sapply(metadata_split, `[`, 4)

meta_df <- data.frame(
  ID = sapply(metadata_split, `[`, 1),
  Species = sapply(metadata_split, `[`, 2),
  Condition = sapply(metadata_split, `[`, 3),
  Location = sapply(metadata_split, `[`, 4),
  Abbrevations = sapply(split_abbreviations, "[", 1)
)
rownames(meta_df) <- NULL

# filter out important data:

# sample data:
samples <- data.frame(row.names=meta_df$Abbrevations, Condition=meta_df$Condition, Species=meta_df$Species)
# sample gene count data:
count_df <- data[, -(2:6)]

# Rename the sample columns of 'count_df' using the sample names
colnames(count_df)[-(1)] <- meta_df$Abbrevations
```
## meta_df - dataframe of the metadata of the samples:
```{r}
meta_df
```
## count_df - dataframe of the samples and their counts:
```{r}
count_df
```
## Separate dataframe of samples under Control condition:
```{r}
# Identify columns containing "D"
all_drought_samples <- grep("D", names(count_df), value = TRUE)

# Remove the specified columns from count_df
count_df_C <- count_df[, !colnames(count_df) %in% all_drought_samples]

meta_df_C <- meta_df[!meta_df$Abbrevations %in% all_drought_samples, ]

samples_C <- data.frame(row.names=meta_df_C$Abbrevations, Condition=meta_df_C$Condition, Species=meta_df_C$Species)
print(meta_df_C)
print(samples_C)
```
## Separate dataframe of samples under Drought condition:

```{r}
# Identify columns containing "C"
all_control_samples <- grep("C", names(count_df), value = TRUE)

# Remove the specified columns from count_df
count_df_D <- count_df[, !colnames(count_df) %in% all_control_samples]

meta_df_D <- meta_df[!meta_df$Abbrevations %in% all_control_samples, ]

samples_D <- data.frame(row.names=meta_df_D$Abbrevations, Condition=meta_df_D$Condition, Species=meta_df_D$Species)
print(meta_df_D)
print(samples_D)
```
## Separate dataframe of only N.excelsior samples:
```{r}
# Identify columns containing "E"
all_control_samples <- grep("T", names(count_df), value = TRUE)

# Remove the specified columns from count_df
count_df_E <- count_df[, !colnames(count_df) %in% all_control_samples]

meta_df_E <- meta_df[!meta_df$Abbrevations %in% all_control_samples, ]

samples_E <- data.frame(row.names=meta_df_E$Abbrevations, Condition=meta_df_E$Condition, Species=meta_df_E$Species)
print(meta_df_E)
print(samples_E)
```

## Separate dataframe of only N.truncata samples:
```{r}
# Identify columns containing "E"
all_control_samples <- grep("E", names(count_df), value = TRUE)

# Remove the specified columns from count_df
count_df_T <- count_df[, !colnames(count_df) %in% all_control_samples]

meta_df_T <- meta_df[!meta_df$Abbrevations %in% all_control_samples, ]

samples_T <- data.frame(row.names=meta_df_T$Abbrevations, Condition=meta_df_T$Condition, Species=meta_df_T$Species)
print(meta_df_T)
print(samples_T)
```

# Dataframe of samples without outliers, detected by PCA (see other R notebook: TRU_EXC_EdgeR_DGEanalysis_notebook.rmd)
```{r}
# After PCA inspection, remove samples:
#removesamples <- c("ECR71", "ECR65", "EDR61", "TCR92", "TDR95")
removesamples <- c("TCR92")
# Remove the specified columns from count_df
count_df_outr <- count_df[, !colnames(count_df) %in% removesamples]

meta_df_outr <- meta_df[!meta_df$Abbrevations %in% removesamples, ]

samples_outr <- data.frame(row.names=meta_df_outr$Abbrevations, Condition=meta_df_outr$Condition, Species=meta_df_outr$Species)
print(meta_df_outr)
print(samples_outr)
```
## # Dataframe of N.excelsior samples without outliers,
```{r}
# Identify columns containing "T"
all_control_samples <- grep("T", names(count_df_outr), value = TRUE)

# Remove the specified columns from count_df
count_df_E_outr <- count_df_outr[, !colnames(count_df_outr) %in% all_control_samples]

meta_df_E_outr <- meta_df_outr[!meta_df_outr$Abbrevations %in% all_control_samples, ]

samples_E_outr <- data.frame(row.names=meta_df_E_outr$Abbrevations, Condition=meta_df_E_outr$Condition, Species=meta_df_E_outr$Species)
print(meta_df_E_outr)
print(samples_E_outr)
```
# Dataframe of N.truncata samples without outliers,
```{r}
# Identify columns containing "E"
all_control_samples <- grep("E", names(count_df_outr), value = TRUE)

# Remove the specified columns from count_df
count_df_T_outr <- count_df_outr[, !colnames(count_df_outr) %in% all_control_samples]

meta_df_T_outr <- meta_df_outr[!meta_df_outr$Abbrevations %in% all_control_samples, ]

samples_T_outr <- data.frame(row.names=meta_df_T_outr$Abbrevations, Condition=meta_df_T_outr$Condition, Species=meta_df_T_outr$Species)
print(meta_df_T_outr)
print(samples_T_outr)
```

# DGE Analysis
```{r}
calculate_num_significant_genes <- function(count_matrix, cpm_threshold, FDR = 0.05, lfc_threshold = 1.5) {
  print(cpm_threshold)
  # Filter genes with mean CPM > threshold
  gene_cpm_filter <- apply(cpm(count_matrix), 1, function(x) !(mean(x) < cpm_threshold))
  count_matrix_filtered <- count_matrix[gene_cpm_filter,]
  size_count_matrix_filtered <- dim(count_matrix_filtered)
  num_rows <- size_count_matrix_filtered[1]
  #gene_keep_criteria <- rowSums(cpm(d) > 10) >= 6
  #d_T_outr_high_cpm_filtered <- d_T_outr[gene_keep_criteria,]
  
  # Normalize data by TMM
  normalized_data <- calcNormFactors(count_matrix_filtered, method = "TMM")
  
  # Design matrix for GLM
  design_mat <- model.matrix(~ 0 + count_matrix$samples$group)
  colnames(design.mat) <- levels(count_matrix$samples$group)

  # GLM testing for differential expression
  d_disp <- estimateDisp(normalized_data, robust = T)
  d_GLM <- estimateGLMTagwiseDisp(d_disp, design_mat)
  
  d_fit <- glmFit(d_GLM, design_mat)
  lrtdc <- glmLRT(d_fit, contrast = c(-1, 1))  # Compare groups: 1-control and -1-drought
  
  # Get the number of differentially expressed genes (at FDR < 0.05 and log2-fold-change > threshold)
  deGLM <- decideTestsDGE(lrtdc, adjust.method = "BH", p.value = FDR, lfc = lfc_threshold)
  significant_genes <- rownames(count_matrix_filtered)[as.logical(deGLM)]
  num_significant_genes <- length(significant_genes)
  print(num_rows)
  return(list(num_rows, num_significant_genes))
}

plot_significant_genes_vs_cpm_0to1 <- function(count_matrix, num_steps = 10, FDR = 0.05, lfc_threshold = 1.5) {
  cpm_values <- seq(0, 1, length.out = num_steps)
  df <- data.frame(CPM_Threshold = cpm_values, Significant_Genes = numeric(length(cpm_values)), Num_Rows = numeric(length(cpm_values)))
  
  for (i in 1:length(cpm_values)) {
    cpm_threshold <- cpm_values[i]
    calculation_output <- calculate_num_significant_genes(count_matrix, cpm_threshold, FDR, lfc_threshold)
    df[i, "Num_Rows"] <- calculation_output[[1]]  
    df[i, "Significant_Genes"] <- calculation_output[[2]]
  }
  
  sign_gene_threshold_plot <- ggplot(df, aes(x = CPM_Threshold, y = Significant_Genes)) +
    geom_line() +
    geom_point() +
    labs(x = "CPM Threshold", y = "Number of Significant Genes") +
    theme_minimal()
  
  return(list(df, sign_gene_threshold_plot))
}

plot_significant_genes_vs_cpm <- function(count_matrix, max_cpm = 10, FDR = 0.05, lfc_threshold = 1.5) {
  cpm_values <- 0:max_cpm
  df <- data.frame(CPM_Threshold = cpm_values, Significant_Genes = numeric(length(cpm_values)), Num_Rows = numeric(length(cpm_values)))
  
  for (i in 1:length(cpm_values)) {
    cpm_threshold <- cpm_values[i]
    calculation_output <- calculate_num_significant_genes(count_matrix, cpm_threshold, FDR, lfc_threshold)
    df[i, "Num_Rows"] <- calculation_output[[1]]  
    df[i, "Significant_Genes"] <- calculation_output[[2]]
  }
  
  sign_gene_threshold_plot <- ggplot(df, aes(x = CPM_Threshold, y = Significant_Genes)) +
    geom_line() +
    geom_point() +
    labs(x = "CPM Threshold", y = "Number of Significant Genes") +
    theme_minimal()
  
  return(list(df, sign_gene_threshold_plot))
}
```

## DGE analysis of N.excelsior with DeSeq2
```{r}
# Create DESeqDataSet
dds <- DESeqDataSetFromMatrix(countData = count_df_E_outr[-1], colData = samples_E_outr, design = ~Condition)
```
## Pre-filter samples with low count for better visualization and faster processing (not necessary, DESeq2 has independent filtering)
```{r}
#smallestGroupSize <- 3
#keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
#dds <- dds[keep,]
```

```{r}
dds <- DESeq(dds) # DESeq awaits un-normalized counts for its statistical model to hold. DESeq2 corrects for library size internally.
resultsNames(dds) # lists the coefficients
res <- results(dds, name="Condition_drought_vs_control")
res <- results(dds, contrast=c("Condition","drought","control"))
geneids = count_df$Geneid
rownames(res) <- geneids


# Mapping function: ID to index
id2index <- function(id) {
  if (id %in% rownames(res)) {
    return(which(rownames(res) == id))
  } else {
    return(NA)  # Return NA if the ID is not found
  }
}

# Mapping function: Index to ID
index2id <- function(index) {
  if (index >= 1 && index <= nrow(res)) {
    return(rownames(res)[index])
  } else {
    return(NA)  # Return NA if the index is out of range
  }
}
```

## Show DGE Result of N. excelsior:
```{r}
res
```
## Show DGE Result of N. excelsior, with Control association log2FC values:
```{r}
res_C_shrink <- lfcShrink(dds, coef="Condition_drought_vs_control", type="apeglm")
res_C_shrink    # adjust FoldChange with association to Control
```
## Order by p-value ascending
```{r}
resOrdered <- res[order(res$pvalue),]
resOrdered
```
```{r}
summary(res)
```
## genes count with p-values less than 0.1
```{r}
sum(res$padj < 0.1, na.rm=TRUE)
```
## Filtering:

### Plots of independent filterings:
### Distribution of p-value over mean of normalized counts:
```{r}
# Calculate baseMean + 1 and -log10(p-value) for filtered and non-filtered genes
filtered_baseMean <- res$baseMean[!use] + 1
filtered_neglog_pvalue <- -log10(res$pvalue[!use])

nonfiltered_baseMean <- res$baseMean[use] + 1
nonfiltered_neglog_pvalue <- -log10(res$pvalue[use])

# Create the plot
plot(res$baseMean + 1, -log10(res$pvalue),
     log = "x", xlab = "Mean of normalized counts",
     ylab = expression(-log[10](pvalue)),
     ylim = c(0, 30),
     cex = 0.4, col = rgb(0, 0, 0, 0))

# Mark points for filtered genes
points(filtered_baseMean, filtered_neglog_pvalue, col = "red", pch = 15, cex = 0.3)

# Mark points for non-filtered genes
points(nonfiltered_baseMean, nonfiltered_neglog_pvalue, col = "black", pch = 15, cex = 0.3)

# Add legend
legend("topright", legend = c("Filtered genes", "Non-filtered genes"), 
       col = c("red", "black"), pch = 16)

```
### Histogram of filtered samples:
```{r}
use <- res$baseMean > metadata(res)$filterThreshold
h1 <- hist(res$pvalue[!use], breaks=0:50/50, plot=FALSE)
h2 <- hist(res$pvalue[use], breaks=0:50/50, plot=FALSE)
colori <- c(`do not pass`="khaki", `pass`="powderblue")
barplot(height = rbind(h1$counts, h2$counts), beside = FALSE,
        col = colori, space = 0, main = "", ylab="frequency", xlab = "p - value")
text(x = c(0, length(h1$counts)), y = 0, label = paste(c(0,1)),
     adj = c(0.5,1.7), xpd=NA)
legend("topright", fill=rev(colori), legend=rev(names(colori)))
```
### Cook - Outlierdetection
```{r}
par(mar=c(8,5,2,2))
boxplot(log10(assays(dds)[["cooks"]]), range=0, las=2)
```
```{r}
W <- res$stat
maxCooks <- apply(assays(dds)[["cooks"]],1,max)
idx <- !is.na(W)
plot(rank(W[idx]), maxCooks[idx], xlab="rank of Wald statistic", 
     ylab="maximum Cook's distance per gene",
     ylim=c(0,5), cex=.4, col=rgb(0,0,0,.3))
m <- ncol(dds)
p <- 3
abline(h=qf(.99, p, m - p))

# Plot the line for the threshold value
abline(h = metadata(res)$filterThreshold[[1]], col = "red", lty = 2)
```

### The default is to use the 99% quantile of the F(p,m-p) distribution (with p the number of parameters including the intercept and m number of samples)

### Plot of filter statistic function (FDR vs quantile of mean normalized count)
```{r}
plot(metadata(res)$filterNumRej, 
     type="b", ylab="number of rejections",
     xlab="quantiles of filter")
lines(metadata(res)$lo.fit, col="red")
abline(v=metadata(res)$filterTheta)
```

### The results function maximizes the number of rejections (adjusted p value less than a significance level), over the quantiles of a filter statistic (the mean of normalized counts). The threshold chosen (vertical line) is the lowest quantile of the filter for which the number of rejections is within 1 residual standard deviation to the peak of a curve fit to the number of rejections over the filter quantiles:

# Inspect automatic Filter of DESeq2:
```{r}
metadata(res)$filterThreshold
```
```{r}
print(metadata(res)$alpha)
```
```{r}
resNoFilt <- results(dds, independentFiltering=FALSE)
addmargins(table(filtering=(res$padj < .1),
                 noFiltering=(resNoFilt$padj < .1)))
```

### 29224 genes are not significant in both. 564 genes are significant (FDR < 0.1), but were filtered. 7683 non-filtered genes have FDR < 0.1.

### Summary of Filtered Results:
```{r}
summary(res)
```
```{r}
sum(res$padj < 0.1, na.rm=TRUE)
```


### Further Filtering? by FDR = 0.05. Not Necessary, as Wald plotting also chooses low alpha fo significance.
```{r}
# Get the indices of rows with padj less than 0.05 and not NA
alpha05_filter_indices <- which(!is.na(res$padj) & res$padj < 0.05)

# Filter res based on the indices
res05 <- res[alpha05_filter_indices, ]

# Summary of results
summary(res05)
```
### Plot Dispersion plot: Blue are kept genes, black were filtered.
```{r}
plotDispEsts(dds)
```


```{r}
sum(res05$padj < 0.05, na.rm=TRUE)
```
### log2FoldChange - Significance Tests with Wald tests with logFC > 1.5:
```{r}
# Shrink data and prepare them for plotMA:
lfc_threshold = 1.5
res_C_shrink_lFC2 <- lfcShrink(dds, coef=2, type="apeglm", lfcThreshold=lfc_threshold)
rownames(res_C_shrink_lFC2) <- geneids

# res_C_shrink_lFC2_alpha05 = res_C_shrink_lFC2[alpha05_filter_indices, ] # found to be not necessary
```
### Click all significant genes (blue points) of plotMA. Copy+Paste and run plotMA in console to interact with it by running idx. Left click for getting index, Right click to retrieve indices.
```{r}
ylim <- c(-12,12)
drawLines <- function() abline(h=c(-lfc_threshold, lfc_threshold),col="dodgerblue",lwd=2)
plotMA(res_C_shrink_lFC2, ylim=ylim); drawLines()
#dev.off()
idx <- identify(res_C_shrink_lFC2$baseMean, res_C_shrink_lFC2$log2FoldChange)
```
### Get all genes of the clicked indices:
```{r}
identified_genes <- rownames(res)[idx]
identified_genes
identified_indices <- sapply(identified_genes, id2index)
identified_indices
```
### Alternatively... just filter all genes by the given alpha 0.005 for which colouring of the points occurs.
```{r}
DEGdata <- plotMA(res_C_shrink_lFC2, ylim=ylim, returnData=TRUE)

# Filter rows based on conditions, handling NA values
filtered_rows <- as.integer(rownames(DEGdata[DEGdata$isDE == TRUE, ]))
res_C_shrink_lFC2
# Get the rownames of the filtered rows
filtered_gene_ids <- sapply(filtered_rows, index2id)
filtered_indices <- sapply(filtered_gene_ids, id2index)
filtered_gene_ids
filtered_indices
```

# Size of filtered diff. expr genes
```{r}
length(filtered_gene_ids)
```


### Comparing both methods to retrieve gene names:
```{r}
# Find overlaps using intersect()
overlapping_geneids <- intersect(filtered_gene_ids, identified_genes)
overlapping_indices <- sapply(overlapping_geneids, id2index)
# Calculate the percentage
percentage <- (length(overlapping_geneids) / length(identified_genes)) * 100

# Print overlapping values
print(overlapping_indices)
```

```{r}
# Print the percentage
print(percentage)
```
```{r}
plotCounts(dds, gene=1002, intgroup="Condition")

```
```{r}

# Restructure Results for Saving:
save_df_EXC <- res_C_shrink
rownames(save_df_EXC) <- 1:nrow(save_df_EXC)
#EXC_gene_indices <- sapply(EXC_Gene_ID, id2index) # If there is a subset of the results

save_df_EXC
# Add columns: svalue, Regulation, GeneID
# change rownames to indices

save_df_EXC$svalue <- res_C_shrink_lFC2$svalue

save_df_EXC$Regulation <- ifelse(save_df_EXC$log2FoldChange > 0, "Upregulated", "Downregulated")

save_df_EXC$Gene_ID <- geneids

save_df_EXC
colnames(save_df_EXC) <- c("BaseMeans", "logFC", "lfcSE", "PValue", "FDR", "svalue", "Regulation", "Gene_ID")
# Reorder by Fold-Change:
save_df_EXC <- save_df_EXC[order(save_df_EXC$logFC, na.last = NA), ]

# Remove filtered samples:
save_df_EXC <- save_df_EXC[complete.cases(save_df_EXC) , ]

#specify significant genes: (from plot: svalue = 0.005 and FC = 1.5) or (FDR < 0.05)
significant_genes <- save_df_EXC[save_df_EXC$FDR < 0.05, ]
significant_genes
write.csv(significant_genes, "DESeq2_DEgenes_EXC_Root.csv")
```

```{r}
significant_genes
```


```{r}

```
# GO retrieval:
```{r}
pacman::p_load("BiocManager", "topGO", "stringr", "GOplot")
library("topGO")
library("GOplot")
library("xtable")
#library("drake")

topGoAnalysis <- function(ontology, allGenes, gene2GO, nbNodes){
  GOdata <- new("topGOdata", ontology = ontology, allGenes = allGenes, annot = annFUN.gene2GO, gene2GO = gene2GO)
  resultWeight01 <- runTest(GOdata, statistic = "fisher")
  allRes <- GenTable(GOdata, weight01_pval=resultWeight01, orderBy = "weight01", ranksOf = "weight01",topNodes = nbNodes)
  allRes <- cbind(allRes,"BP")
  colnames(allRes) <- c("GO.ID","Term","Annotated","Significant","Expected","weight01_pval", "branch")
  return(allRes)
}

change_names <- function(data, name_list){
  colnames(data) <- name_list
  return(data)
}

rename <- function(table, geneNames){
  names(table) <- geneNames
  return(table)
}

attach_enriched_go_genes <- function(enriched_go_with_my_genes){
  enriched_go_with_my_genes.list = c()
  for (i in 1:length(enriched_go_with_my_genes)){
    enriched_go_with_my_genes.list = c(enriched_go_with_my_genes.list, enriched_go_with_my_genes[[i]])
  }
  return(enriched_go_with_my_genes.list)
}


circle_dat <- function(terms, genes){
  # Data Formatting
  colnames(terms) <- tolower(colnames(terms))
  terms$genes <- toupper(terms$genes)
  genes$ID <- toupper(genes$ID)
  
  # Parse Gene Lists
  tgenes <- strsplit(as.vector(terms$genes), ', ')
  count <- sapply(tgenes, length)
  
  # Extract LogFC Values
  if (length(tgenes[[1]]) == 1) tgenes <- strsplit(as.vector(terms$genes), ',')
  count <- sapply(1:length(tgenes), function(x) length(tgenes[[x]]))
  logFC <- sapply(unlist(tgenes), function(x) {
  match_indices <- match(x, genes$ID)
  if (any(is.na(match_indices))) {
    # Handle NA values (no match found)
    numeric_values <- rep(0, length(match_indices))
    numeric_values[!is.na(match_indices)] <- genes$logFC[match_indices[!is.na(match_indices)]]
    numeric_values
  } else {
    genes$logFC[match_indices]
  }
})
  if(class(logFC) == 'factor'){
    logFC <- gsub(",", ".", gsub("\\.", "", logFC))
    logFC <- as.numeric(logFC)
  }
  
  # Calculate Z-score
  s <- 1; zsc <- c()
  for (c in seq_along(count)) {
    e <- s + count[c] - 1
    value <- logFC[s:e]
    zsc <- c(zsc, sum(value, na.rm = FALSE) / sqrt(count[c]))
    s <- e + 1
  }
  
  # Create Output Data Frame
  df <- data.frame(
    category = rep(as.character(terms$category), count),
    term = rep(as.character(terms$term), count),
    count = rep(count, count),
    genes = as.character(unlist(tgenes)),
    logFC = logFC,
    adj_pval = rep(terms$adj_pval, count),
    zscore = rep(zsc, count),
    stringsAsFactors = FALSE
  )
  
  return(df)
}


plot_GO <- function(data_GO,loc){
  #read in data matrix
  data <- read.table(data_GO, sep = "\t")
  colnames(data) <- c("category","ID","term","count","adj_pval","zscore")
  # define colors of bars by zscores
  ggplot(data, aes(x=reorder(term,zscore),y=-log10(adj_pval)))+ #x=reorder(term,zscore),y=-log10(adj_pval)
    #labs(title=paste0("Loc ",loc)) +
    geom_col(aes(fill = zscore)) +
    scale_fill_gradient(low="blueviolet",high="goldenrod1", 
                        labels=c(paste0("Underexpressed","\n","in L"),
                                 paste0("Overexpressed","\n", "in L")),
                        breaks=c(min(data$zscore),max(data$zscore)),
                        limits=c(min(data$zscore),max(data$zscore))) +
    geom_text(aes(label=count), 
              hjust=-1, size=5) +
    xlab("")+
    theme_bw()+
    coord_flip(ylim = c(0, 5.3)) +
    #scale_y_reverse() +
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank())+
    #coord_cartesian(ylim = c(0, 4.6))+
    theme(axis.text.x = element_text(
      angle = 0,
      hjust = 0.3,
      vjust = 0.5))+
    theme(text=element_text(family="Arial"))+
    theme(axis.text=element_text(size=15), axis.text.y = element_text(size=15)) +
    #       axis.title=element_text(size=14,face="bold"))
    theme(legend.text=element_text(size=13),
          legend.title = element_text(size=15, vjust=2.1),
          axis.title.x = element_text(size=15))
}

#Make barplot of up and down-regulated functions
GOBar <- function(data, display, order.by.zscore = T, title, zsc.col, contrast){
  id <- adj_pval <- zscore <- NULL
  colnames(data) <- tolower(colnames(data))
  data$adj_pval <- -log(data$adj_pval, 10)
  sub <- data[!duplicated(data$term), ]
  sub <- sub[order(sub$zscore, decreasing = T), ]
  print(min(sub$zscore))
  print(max(sub$zscore))
  g <-  ggplot(sub, aes(y = factor(term, levels = rev(unique(stats::reorder(term, adj_pval)))), x = adj_pval, fill = zscore)) +
    geom_bar(stat = 'identity', position=position_dodge()) +
    theme_bw() +
    scale_fill_gradient2('Log(FC)\n(z-score)\n', space = 'Lab',
                     guide = guide_colourbar(title.position = "top", title.hjust = 0),
                     breaks = c(min(sub$zscore), max(sub$zscore)),
                     low = "blue", mid = "white", high = "red") +
    scale_color_manual(values=c("black", "black", "black"), guide = "none") +
    labs(title = '', y = '', x = '-log (adj p-value)') +
    geom_text(aes(y=term, x=-0.4, label=count, color = "black")) + theme(axis.text.y = element_text(size = 12))
  g
}
```

```{r}
#Load data
geneID2GO <- readMappings(file = "/Users/daviddornig/Documents/botany_project/data/nblab360_annotation_topGO.csv", IDsep = ";")
GO2geneID <- inverseList(geneID2GO)
geneNames <- names(geneID2GO)
deGenes_EXC_DESeq2 <- read.csv("/Users/daviddornig/Documents/botany_project/data/TRU_EXC_Roots/EXC/DESeq2_DEgenes_EXC_Root.csv", row.names = 1)
```

```{r}

analyze_DE_genes_GO_Terms <- function(deGenes) {
  # Assuming you have other necessary objects and functions defined
  
  name_list = c("GO.ID","Term","Annotated","Significant","Expected","weight01_pval", "branch")
  table = as.factor(geneNames) %in% deGenes$Gene_ID
  int_table = as.integer(table)
  int_fac_table = factor(int_table)
  fac_table = rename(table = int_fac_table, geneNames = geneNames)

  GOdata_BP = new("topGOdata", ontology = "BP", allGenes = fac_table, annot = annFUN.gene2GO, gene2GO = geneID2GO)
  GOdata_MF = new("topGOdata", ontology = "MF", allGenes = fac_table, annot = annFUN.gene2GO, gene2GO = geneID2GO)
  GOdata_CC = new("topGOdata", ontology = "CC", allGenes = fac_table, annot = annFUN.gene2GO, gene2GO = geneID2GO)
  resultWeight01_BP = runTest(GOdata_BP, statistic = "fisher")
  resultWeight01_MF = runTest(GOdata_MF, statistic = "fisher")
  resultWeight01_CC = runTest(GOdata_CC, statistic = "fisher")

  allRes_BP1 = GenTable(GOdata_BP, weight01_pval=resultWeight01_BP, orderBy = "weight01", ranksOf = "weight01",topNodes = 4888)
  allRes_MF1 = GenTable(GOdata_MF, weight01_pval=resultWeight01_MF, orderBy = "weight01", ranksOf = "weight01",topNodes = 2436)
  allRes_CC1 = GenTable(GOdata_CC, weight01_pval=resultWeight01_CC, orderBy = "weight01", ranksOf = "weight01",topNodes = 929)
 return(list(list(GOdata_BP, resultWeight01_BP), list(GOdata_MF, resultWeight01_MF), list(GOdata_CC, resultWeight01_CC), list(allRes_BP1, allRes_MF1, allRes_CC1)))
}
``` 

```{r}
structure_DE_genes_GO_data <- function(GOdata_BP, GOdata_MF, GOdata_CC, allRes_BP1, allRes_MF1, allRes_CC1, FDR=0.05){
  # Structure Dataframes:
  allRes_BP2 = cbind(allRes_BP1,"BP")
  allRes_MF2 = cbind(allRes_MF1,"MF")
  allRes_CC2 = cbind(allRes_CC1,"CC")
  
  allRes_BP = change_names(data = allRes_BP2, name_list = name_list)
  allRes_MF = change_names(data = allRes_MF2, name_list = name_list)
  allRes_CC = change_names(data = allRes_CC2, name_list = name_list)
  
  allRes1 = rbind(allRes_BP, allRes_MF)
  allRes = rbind(allRes1, allRes_CC)

  allGO = c(genesInTerm(GOdata_BP), genesInTerm(GOdata_MF), genesInTerm(GOdata_CC))

  # Create final table
  SAM_ANOTATION = lapply(allGO, function(x) x[x %in% deGenes$Gene_ID])
  enriched_go_with_my_genes = lapply(SAM_ANOTATION[allRes[, 1]], paste0, collapse = ", ")
  enriched_go_with_my_genes.list = attach_enriched_go_genes(enriched_go_with_my_genes)
  go_dataframe = data.frame("Category" = allRes$branch, "ID" = allRes$GO.ID, "Term" = allRes$Term, "Genes" = as.vector(enriched_go_with_my_genes.list), "adj_pval" = as.numeric(sub(",", ".", allRes$weight01_pval, fixed = TRUE)))

  DE_genes = deGenes
  EC_genelist = data.frame("ID" = DE_genes$Gene_ID, "logFC" = DE_genes$logFC, "logCPM" = DE_genes$logCPM, "P.Value" = DE_genes$PValue)
  circ = circle_dat(go_dataframe, EC_genelist)
  circ_significant <- circ[circ$adj_pval <= FDR,]
  reduced_circ <- reduce_overlap(circ_significant, overlap = 0.6)
  
  # Return the final tables and plot
  return(list(go_dataframe = go_dataframe, reduced_circ, EC_genelist,  circ, circ_significant))
}  

plot_all_GO_data <- function(reduced_circ, 
                             Method, species, contrast, background, 
                             savepath, 
                             CPM, FDR, logFC){
  contrast = "DroughtvsControl"
  
  plot_GO_BP <- GOBar(subset(reduced_circ, category == 'BP'), contrast = contrast)
  plot_GO_MF <- GOBar(subset(reduced_circ, category == 'MF'), contrast = contrast)
  plot_GO_CC <- GOBar(subset(reduced_circ, category == 'CC'), contrast = contrast)
  # Save the plot to a PDF file
  setwd(savepath)
  ggsave(paste0(Method,"_", "Barplot_TopGOBP_LogFC_Zscore.", species, "_", contrast,"_", background, "_at","CPM", CPM, "_FDR", FDR,"_FC", logFC, ".pdf"), plot_GO_BP, width = 12, height = 6)
  ggsave(paste0(Method,"_", "Barplot_TopGOMF_LogFC_Zscore.", species, "_", contrast,"_", background, "_atCPM", CPM, "_FDR", FDR,"_FC", logFC, ".pdf"), plot_GO_MF, width = 12, height = 6)
  ggsave(paste0(Method,"_", "Barplot_TopGOCC_LogFC_Zscore.", species, "_", contrast,"_", background, "_atCPM", CPM, "_FDR", FDR,"_FC", logFC, ".pdf"), plot_GO_CC, width = 12, height = 6)

  # Return the final tables and plot
  return(list(plot_GO_BP, plot_GO_MF, plot_GO_CC))
}

create_GO_network_graphs <- function(GO_data_BP, resultWeight_BP, GO_data_MF, resultWeight_MF,
                                     GO_data_CC, resultWeight_CC, 
                                     Method, species, contrast, background, 
                                     Nodes = 10, 
                                     savepath,
                                     CPM, FDR, logFC){
  setwd(savepath)
  printGraph(GOdata_BP, resultWeight_BP, Nodes, fn.prefix = paste0(Method,"_","GO_Network_BP", "_", species, "_",
                                                                                contrast, "_", background, "_atCPM", CPM, "_FDR", FDR,"_FC", logFC, "_N", Nodes), useInfo = "all", pdfSW = TRUE)
  
  printGraph(GOdata_MF, resultWeight_MF, Nodes, fn.prefix = paste0(Method,"_","GO_Network_MF", "_", species, "_",
                                                                                contrast, "_", background, "_atCPM", CPM, "_FDR", FDR,"_FC", logFC, "_N", Nodes), useInfo = "all", pdfSW = TRUE)
  
  printGraph(GOdata_CC, resultWeight_CC, Nodes, fn.prefix = paste0(Method,"_","GO_Network_CC", "_", species, "_",
                                                                                contrast, "_", background, "_atCPM", CPM, "_FDR", FDR,"_FC", logFC, "_N", Nodes), useInfo = "all", pdfSW = TRUE)
  
}
```

```{r}
output1_EXC_DESeq2 = analyze_DE_genes_GO_Terms(deGenes_EXC_DESeq2)

output2_EXC_DESeq2 = structure_DE_genes_GO_data(output1_EXC_DESeq2[[1]][[1]], output1_EXC_DESeq2[[2]][[1]], output1_EXC_DESeq2[[3]][[1]],
                                               output1_EXC_DESeq2[[4]][[1]], output1_EXC_DESeq2[[4]][[2]], output1_EXC_DESeq2[[4]][[3]],
                                                FDR=0.05)

go_dataframe_EXC_DESeq2 <- output2_EXC_DESeq2[[1]]

Method = "DESeq2"
species = "ER"
contrast = "DvC"
background = "92_outlier"
savepath <- "/Users/daviddornig/Documents/botany_project/Images/DGE/DESeq2"
go_plots_EXC_DESeq2 <- plot_all_GO_data(output2_EXC_DESeq2[[2]], Method, species, contrast, background, savepath, "Mean1.8", 0.05, 1.5)

create_GO_network_graphs(output1_EXC_DESeq2[[1]][[1]], output1_EXC_DESeq2[[1]][[2]], output1_EXC_DESeq2[[2]][[1]],
                         output1_EXC_DESeq2[[2]][[2]], output1_EXC_DESeq2[[3]][[1]], output1_EXC_DESeq2[[3]][[2]],
                         Method, species, contrast, background, Nodes=3, savepath, "Mean1.8", 0.05, 1.5)

```
```{r}
go_plots_EXC_DESeq2
```
```{r}
df = go_dataframe_EXC_DESeq2
df2 <-  output2_EXC_DESeq2[[2]]
df2
water_deprivation_row <- df[grepl("water deprivation", df$Term, ignore.case = TRUE), ]
zscore <- df2[grepl("water deprivation", df2$term, ignore.case = TRUE), ]
zscore
# Print the resulting row
print(water_deprivation_row)
```
```{r}
# Extract gene names under the $Genes column
gene_names <- unlist(strsplit(water_deprivation_row$Genes, ", "))
gene_names
# Call id2index function to get indices
gene_indices <- sapply(gene_names, id2index)
gene_indices
```
```{r}
dfBP <-  output1_EXC_DESeq2[[4]][[1]]
dfBP_sign <- output2_EXC_DESeq2[[2]]
dfBP_sign
metabolic_row <- dfBP_sign[grepl("process", dfBP_sign$term, ignore.case = TRUE), ]
metabolic_zscore <- dfBP[grepl("metabolic", dfBP$term, ignore.case = TRUE), ]
metabolic_row
metabolic_zscore

 plot_GO_BP <- GOBar(subset(dfBP_sign, category == 'BP'), contrast = contrast)
 plot_GO_BP
```


```{r}
dds_EXC <- DESeqDataSetFromMatrix(countData = count_df_E_outr[-1], colData = samples_E_outr, design = ~Condition)
dds_EXC <- DESeq(dds_EXC)
```

```{r}
plotCounts(dds_EXC, gene=2679, intgroup="Condition")

```


